Reanalysis Step 2: Multiple Imputation

### 2025-08-30 reanalysis-2-multiple_imputation.qmd
perfect prediction

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = dplyr::row_number(),
    # Store cohort_id separately for later use (not in imputation)
    cohort_id_original = cohort_id
  ) |>
  dplyr::select(row_id, cohort_id_original, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2),
                       labels = c("No_DR", "NPDR", "PDR"),
                       ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  ) |>
  # Fellow-eye variables (computed within patient clusters)
  dplyr::group_by(e_mrn_deidentified) |>
  dplyr::mutate(
    n_eyes = dplyr::n(),
    eye_index = dplyr::row_number(),
    fellow_baseline_VA = dplyr::if_else(n_eyes == 2, baseline_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_intermediate_VA = dplyr::if_else(n_eyes == 2, intermediate_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_outcome_VA = dplyr::if_else(n_eyes == 2, outcome_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_dr = dplyr::if_else(n_eyes == 2,
                               person_dr[3 - eye_index],
                               factor(NA, levels = levels(person_dr), ordered = TRUE))
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-eye_index)
```

@@

# IDs and fellow-eye variables: never impute
custom_methods["row_id"] <- ""
custom_methods["e_mrn_deidentified"] <- ""
custom_methods["cohort_id_original"] <- ""  # Stored for later, not imputed
for (fv in c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")) {
  if (fv %in% names(custom_methods)) custom_methods[fv] <- ""
}

@@

# Never impute IDs (row = 0) and never use them as predictors (col = 0)
for (idv in c("row_id", "e_mrn_deidentified", "cohort_id_original")) {
  if (idv %in% rownames(pred_matrix)) pred_matrix[idv, ] <- 0
  if (idv %in% colnames(pred_matrix)) pred_matrix[, idv] <- 0
}

@@

# Minimal completeness verification: first completed dataset has no missing values
first_imp <- complete(mice_results, 1)
missing_after <- colSums(is.na(first_imp))
missing_after <- missing_after[missing_after > 0]
if (length(missing_after) > 0) {
  log_message("WARNING: Missing values remain after imputation:")
  print(missing_after)
} else {
  log_message("SUCCESS: All missing data imputed in first completed dataset.")
}
```

@@

# First add back cohort_id from cohort_id_original
imputed_datasets <- lapply(imputed_datasets, function(imp_data) {
  imp_data$cohort_id <- imp_data$cohort_id_original
  imp_data$cohort_id_original <- NULL  # Remove the temporary column
  imp_data
})

>>>>

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = dplyr::row_number()
  ) |>
  dplyr::select(row_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2),
                       labels = c("No_DR", "NPDR", "PDR"),
                       ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  ) |>
  # Fellow-eye variables (computed within patient clusters)
  dplyr::group_by(e_mrn_deidentified) |>
  dplyr::mutate(
    n_eyes = dplyr::n(),
    eye_index = dplyr::row_number(),
    fellow_baseline_VA = dplyr::if_else(n_eyes == 2, baseline_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_intermediate_VA = dplyr::if_else(n_eyes == 2, intermediate_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_outcome_VA = dplyr::if_else(n_eyes == 2, outcome_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_dr = dplyr::if_else(n_eyes == 2,
                               person_dr[3 - eye_index],
                               factor(NA, levels = levels(person_dr), ordered = TRUE))
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-eye_index)
```

@@

# IDs and fellow-eye variables: never impute
custom_methods["row_id"] <- ""
custom_methods["e_mrn_deidentified"] <- ""
for (fv in c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")) {
  if (fv %in% names(custom_methods)) custom_methods[fv] <- ""
}

@@

# Never impute IDs (row = 0) and never use them as predictors (col = 0)
for (idv in c("row_id", "e_mrn_deidentified")) {
  if (idv %in% rownames(pred_matrix)) pred_matrix[idv, ] <- 0
  if (idv %in% colnames(pred_matrix)) pred_matrix[, idv] <- 0
}

@@

# Minimal completeness verification: first completed dataset has no missing values
first_imp <- complete(mice_results, 1)
missing_after <- colSums(is.na(first_imp))

# Check specific key variables
key_vars_check <- c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR", "person_dr")
log_message("\nMissing values in key variables after imputation:")
for (v in key_vars_check) {
  if (v %in% names(first_imp)) {
    n_miss_after <- sum(is.na(first_imp[[v]]))
    n_miss_before <- sum(is.na(df_mice[[v]]))
    log_message(paste0(v, ": Before=", n_miss_before, ", After=", n_miss_after))
  }
}

# Report on variables that still have missing values
missing_after <- missing_after[missing_after > 0]
if (length(missing_after) > 0) {
  log_message("\nWARNING: Variables with remaining missing values:")
  print(missing_after)

  # Check if these are the fellow variables (which shouldn't be imputed)
  fellow_vars_missing <- intersect(names(missing_after), 
                                   c("fellow_baseline_VA", "fellow_intermediate_VA", 
                                     "fellow_outcome_VA", "fellow_dr"))
  if (length(fellow_vars_missing) > 0) {
    log_message("Note: Fellow variables are expected to have missing values (single eyes)")
  }
} else {
  log_message("\nSUCCESS: All missing data imputed in first completed dataset.")
}
```

@@

# Add back cohort_id from the original data
imputed_datasets <- lapply(imputed_datasets, function(imp_data) {
  # Add cohort_id back from original data based on row_id
  imp_data <- imp_data |>
    dplyr::left_join(
      df_original_full |> dplyr::select(row_id, cohort_id),
      by = "row_id"
    )
  imp_data
})
